\glsresetall
\chapter{Epidemiological considerations}
\label{chap:epidemiological_considerations}
\begin{tcolorbox}[title={Contributions of this chapter}]
    
    \paragraph{\nameref{sec:objectives}}

    \paragraph{\nameref{sec:data}}

    \paragraph{\nameref{sec:dessiderata}}

    \paragraph{\nameref{sec:measures_of_epidemic_spread}}

\end{tcolorbox}
\newpage

The spread of infectious diseases, such as \acrshort{c19}, is a complex phenomenon. For \acrshort{c19}, this complexity arises from the interplay of many factors. Studying these influences allows us to define the aims and challenges of epidemiological modeling in the context of this thesis. It also will guide us towards desirable and possible outcomes of our efforts from an applied perspective.

First of all, there is considerable heterogeneity in the way the disease progresses once an individual is infected \citep{Salzberger2021Epidemiology}. Some infectees may show few to no symptoms but are still highly infectious \citep{Byambasuren2020Estimating}, and disease progression is tightly linked to age and preexisting comorbidities \citep{Biswas2020Association}. Additionally, different variants of \acrshort{scov2} differ in key epidemiological characteristics such as the reproduction number \citep{Du2022Reproduction} and mortality \citep{Hughes2023Effect}. 

Second, the spread is highly dependent on the contact behavior in the population, as the infector has to be in close physical proximity to the infectee to infect them. These contact patterns are an essential component of any mathematical model for infectious diseases, as they define how the epidemic evolves. While there are some empirical studies \citep{Tomori2021Individual,Mossong2008Social}, capturing the contact behavior at certain points in time, in the context of an ongoing epidemic these patterns are subject to change, not only in intensity but also in shape \citep{Tomori2021Individual}. As contact restrictions were put into place or lifted, mask-wearing was enforced and home office and schooling became commonplace, so did the number of contacts change and occur under different circumstances. 

Finally, as the virus spread in the population and vaccinations became available, the population became partially immune against the disease, if not against infection \todo{citep}. This immunity affects the spread as well: if an infector has contact with a partially immune individual, the probability of transmission is smaller. Additionally, partial immunity may lead infectors to develop fewer or no symptoms so they may not be aware of being infectious, foregoing quarantine.

Parts of this chapter, especially \Cref{sec:measures_of_epidemic_spread,sec:dessiderata,sec:data}, consist of the ideas published in \citep{Heyder2023Measures}, but have been rewritten to fit better into this thesis.

As statisticians, we are faced with a difficult problem: Which of these factors should we include in our model and how? The answer certainly depends on the epidemiological question under consideration and the availability and quality of data. 

%mathematical epidemiology concerns itself with modelling epidemiolocial systems, from small (local outbreaks) to large (epi/pandemics)
%conclusions from analysis only as good as the model and the data are
%dependening on goal and circumstances different methods are applicable
%by its nature, data are observational so causal claims difficult
%in this thesis I focus on models for larger-scale epidemics, techniques would be flexible enough to deal with smaller scale as well, as long as latent states are gaussian

\input{chapters/02_epidemiology/02_01_objectives.tex}
\input{chapters/02_epidemiology/02_02_measures.tex}
\input{chapters/02_epidemiology/02_03_data.tex}
\input{chapters/02_epidemiology/02_04_dessiderata.tex}